---
title: "Pattern Mining: Market Basket Analysis"
subtitle: ""
author: "Dr. Thiyanga S. Talagala <br/>  Department of Statistics, Faculty of Applied Sciences <br/> University of Sri Jayewardenepura, Sri Lanka"
format: 
  revealjs:
    width: 2000
    height: 1500
    margin: 0.1
    fontsize: 20pt
    theme: [simple, custom.scss]
    transition: slide
    background-transition: fade
    slide-number: true
    show-slide-number: all
    title-slide-attributes: 
      data-background-color: "#081d58"
      data-background-image: none
---

## Market Basket Analysis

- Affinity analysis

- Unsupervised learning

- Frequent itemset mining: To discover which groups of products tend to be purchased together.

---


# Basic concepts


---

## Transaction dataset

:::: {.columns}

::: {.column width="50%"}

```{r, echo=FALSE}
TID <- 1:9
Items <- c("i1, i2, i5", "i2, i4", "i2, i3", "i1, i2, i4", "i1, i3", "i2, i3", "i1, i3", "i1, i2, i3, i5", "i1, i2, i3")
df <- data.frame(TID, Items)
df |> knitr::kable()
```

:::

::: {.column width="50%"}

Item set: Set of items

:::

::::
---

## Suppose we have 100 items. Find the total number of itemsets.

---

## Association rule 

$$ Milk \Rightarrow Bread \text{ [Support = 2%, Confidence = 60%]}$$

- IF (Antecedent)

- THEN (Consequent)

- Support and Confidence measures the strength of association between antecedent and consequent itemset.

---

## Apriori algorithm

Desired support: 20%

Desired confidence: 70%

---

# Step 1:

Translate data into binary incidence matrix format.

---

## Transaction dataset

:::: {.columns}

::: {.column width="30%"}

```{r, echo=FALSE}
TID <- 1:9
Items <- c("i1, i2, i5", "i2, i4", "i2, i3", "i1, i2, i4", "i1, i3", "i2, i3", "i1, i3", "i1, i2, i3, i5", "i1, i2, i3")
df <- data.frame(TID, Items)
df |> knitr::kable()
```

:::

::: {.column width="70%"}



:::

::::

---

## Step 2:

Select itemsets where the minimum support count is 2.


---

## Step 3:

Generate Associate Rules: Compute confidence and lift

---

## Confidence and Lift

In-class demonstration